Influence of Connectivity on Dung Beetle Communities

### 2025-03-03 corridor_docs/eric_ms_thesis/thesis_planning.Rmd
writing

Dung beetles have emerged as a model system with which to test spatial ecology hypotheses and  [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which play an important role in providng ecosystem services for dung removal, secondary seed distribution, and even suppressing populations of parasitic pests[@shepherd_dung_1998; @WOS:000369463100010]. Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

@@

#### messy notes
INTRODUCTION

 These movements are hypothesized to prevent species diversity from declining in fragments, as well as help maintain the ecosystem services provided by these species (at both the patch- and landscape-level)[@burt_ants_2022]. Although there is some evidence that animals disperse between patches via corridors, and that connected patches have higher species diversity than unconnected ones, little work to date has investigated the consequences of these corridor-driven patterns for ecosystem services.

  The removal, breakdown, and burial of animal feces is an important ecosystem service provided by dung beetles such as enhanced nutrient cycling and soil quality and reduction of parasites on methane emissions from dung [@slade_disentangling_2016; @WOS:000350444300013]. Local assemblages of dung beetles can be species-rich with species comprising a broad range of functional traits (e.g., size, foraging style, resource-use) [@WOS:000891747700001]. Previous studies have shown that isolated patches of habitat frequently have lower dung beetle diversity and abundance than areas of continuous habitat, as well as documented their presence in linear strips of habitat that resemble corridors [@WOS:000707546700001]. However, it remains unknown if corridors actually act to reduce the loss of dung beetle species from fragments, if such declines are influenced by inter-specific differences in dispersal capability, and what the consequences of these patterns are for the ecosystems services they provide. One major factor behind this lack of information is the challenge in finding locations where one can assess the role of corridors while also while controlling for confounding factors such as patch size, edge, and corridor length [@haddad_habitat_2015].

 **We sampled the commmunity of of dung beetles at the SRS Corridor Experiment to test the following prediction:** Species Richness, Species Diversity, and Functional Diversity will be higher in patches connected by corridors than in unconnected patches.

- First Paragraph: What is the topic of your introduction and why is it important/interesting/relevant?

- Second Paragraph: What is known about this topic already?

Effects of landscape change are already very well studied, accross different taxa .
effects of edge, patch size, and direct habitat loss.
using corridors to connect fragmented landscapes is also well studied and for many taxa.
dung beetles as a model system for studying movement, functional diversity, and ecosystem health.
dung beetle populations in connected and fragmented habitats .
insect declines in fragmented landscapes.

studying coprophagous insects

- Third Paragraph: What isn't known about this topic and why might it change how we think/act about the topic?

How corridors directly impact dung beetle communitites
are changes driven by dispersal ability

Dung beetles have emerged as a model system with which to test spatial ecology hypotheses and  [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which play an important role in providng ecosystem services for dung removal, secondary seed distribution, and even suppressing populations of parasitic pests[@shepherd_dung_1998; @WOS:000369463100010]. Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

at our study site we can directly compare patch connectivity and patch shape with fragmented landscapes to obtain a strong idea of what landscape features effect dung beetle collection.

- Fourth Paragraph: Why hasn't this thing been studied/assessed/done before?

other work at the corridor project but we are doing dung beetles instead

- Fifth Paragraph: Literally the words "Here we..."

Here sampled the commmunity of of dung beetles at the SRS Corridor Experiment to test the following prediction:

Species richness, diversity, and functional diversity will be higher in patches connected by a movement corridor than in patches that are unconnected.

@@

Biodiversity between patch types was compared using Hill numbers (Jost 2006). We looked at community composition by increasing magnitudes of diversity components (qD) of 0D (species richness), 1D (Shannon entropy), and 2D (Simpson Diversity). Diversity numbers were calculated in R studio using package iNEXT (Chao et al. 2016).

Dung beetles were assigned traits by

@@

# RESULTS
some summary statistics:

Overall, 4995 total dung beetles were collected belonging to 15 species. The most common species were C. vigilans (n=1300), Ateuchus lecontei (n=1112), and Phanaeus igneus (n=919).

>>>>

Dung beetles have emerged as a model system with which to test spatial ecology hypotheses  [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which are well known for driving a multitude of ecosystem functions (Hasan et al. 2024). The removal, breakdown, and burial of animal feces is an important ecosystem service provided by dung beetles such as enhanced nutrient cycling and soil quality and reduction of parasites on methane emissions from dung [@slade_disentangling_2016; @WOS:000350444300013]. Local assemblages of dung beetles can be species-rich with species comprising a broad range of functional traits (e.g., size, foraging style, resource-use) [@WOS:000891747700001]. Previous studies have shown that isolated patches of habitat frequently have lower dung beetle diversity and abundance than areas of continuous habitat, as well as documented their presence in linear strips of habitat that resemble corridors [@WOS:000707546700001].  Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

@@

@@

Biodiversity between patch types was compared using Hill numbers (Jost 2006). We looked at community composition by increasing magnitudes of diversity components (qD) of 0D (species richness), 1D (Shannon entropy), and 2D (Simpson Diversity). Diversity numbers were calculated in R studio using package iNEXT (Chao et al. 2016). Dung beetles were assigned traits by waste removal guild and habitat preference.

@@

# RESULTS

Overall, 5,213 total dung beetles were collected belonging to 16 species. The most common species were C. vigilans (n=1473), Ateuchus lecontei (n=1115), and Phanaeus igneus (n=958). All but one species was found in matrix plots at fifteen species, with the exception of Geotrupes blackburnii, one of our only singletons along with Onthophagous turberculifrons. The winged patch had the next most species at fourteen, rectangular followed with thirteen, and connected patches had the fewest species at twelve (Table 1). Beetle abundance was significantly higher in matrix plots with 33% (n = 1713) of captures followed by 26% (n = 1359) caught in connected patches, 23% (n = 1199) in winged, and lastly 18% (n = 942) from rectanglar patches (Table 2).

Assemblage composition was mostly uniform across patch types, however
